Twenty years of dynamic occupancy models: a review of applications and look towards the future

### 2025-01-23 Manuscript_Initial.qmd
preserve the initial version and begin scaffold for new one

abstract: | 
  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research. Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play. The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates. In the two decades since its introduction, the DOM has become a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.

  In this review we examine how ecologists have applied the DOM in its first two decades of use. Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data were collected for them, and how authors have built their models to achieve their objectives. Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades. The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.

bibliography: references.bib
csl: apa.csl
---

@@

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we have recorded all covariates considered for each model in our sample. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs. We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that users of DOMs have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (37% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models. Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of wetland birds. Several studies also include biotic interactions with other species as covariates, particularly those which focussing on taxa threatened by invasive species. Examples include models with barred owl presence as a covariate on spotted owl occupancy and models with bullfrog covariates on native amphibian occupancy, each of which effectively incorporate species interactions without fitting explicitly multi-species models [@mangan2019; @rowe2019].

::: column-page-inset
```{r covariate table}
#| echo: false
#| message: false
#| label: tbl-covariates
#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories. We calculate the percentage of articles which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;). We also present the percentage of all implementations of each covariate category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate as a non-linear relationship or as part of an interaction with another covariate."

@@

Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate (@tbl-covariates). Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy. These species-specific covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable. For projects interested in making such projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002]. Forty two percent of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@tbl-covariates).

In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function). Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified. Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model. @austin2007 discusses the importance of modelling ecologically realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships. Non-linear responses can be easily accommodated in the DOM by using polynomial transformations and interactions between covariates. Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms (@tbl-covariates). Interactions between covariates were similarly rare, with only 26% of studies considering at least one interaction between terms (@tbl-covariates). The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy. Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008]. This emphasis on more complex responses in SDMs may be due to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear. However, as previously indicated the DOM has also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected. Similar concerns exist for the relatively low level of interactions between covariates: these relationships are commonly expected based on ecological theory, and their exclusion can negatively impact model performance [@guisan2006].

In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter ( @fig-covariates). This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 possible factors.

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates relative to initial occupancy and detection. The lack of any covariates considered for initial occupancy in many studies (37% of articles) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant. Omission of the factors that drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles. Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to the DOM's Markovian nature with implications for the reliability of model outputs. Detection probability is also typically represented with low numbers of environmental covariates (@tbl-covariates). Recall that detectability is dependent on not only observation factors, but also by drivers of abundance and species use; the low number of candidate covariates here relative to colonisation and extinction raises additional questions on whether this variation is always appropriately captured in models.

To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference. Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012]. Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple linear regression may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations. This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested. Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22) frameworks, which differ considerably in their manner of implementation. A summary of modelling practices in DOMs is given in @tbl-modelling.

```{r Modelling summary table}
#| echo: false
#| message: false
#| label: tbl-modelling
#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by frequentist or bayesian implementations. The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review. The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches."

@@

The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection. Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*. Methods used for model selection vary considerably both within and between the two modelling frameworks (@tbl-modelling). For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score. The next most popular method in frequentist studies is to use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one. For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on. The remainder of frequentist studies (9%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM. Across all frequentist implementations, 71% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].

Those Bayesian models which *do* perform model selection tend to take various approaches, with largely idiosyncratic methods across these studies. While direct comparison of model fit is rare in Bayesian methods, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data. Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model. @ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.

Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases. However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020]. Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy models discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate coefficient estimates that may influence model inference. @morin2020's work raises other concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models. The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.

The model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data. Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not yet clear. There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages. @kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model. In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014]. @broms2016 discusses possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM. As in other hierarchical models, model evaluation for DOMs can in general can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known. Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.

Perhaps as a result of these difficulties, assessment of model fit and model performance was rare amongst the articles in our sample. Only 19% of studies tested for goodness-of-fit, and just 10% calculated predictive performance. These rates are considerably lower than those reported for SDMs; for which closer to 50% of articles were found to test both fit and performances [@araújo2019].

@@

Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives). These objective categories were non-exclusive, and many studies fell into multiple categories. Across the study period the DOM has been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions. Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.

Many articles in our sample used the DOM seeking to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives). These studies include @ahumada2013's paper assessing populations trends of 13 forest mammals and @scott2015's assessment of the conservation status of the endangered flattened musk turtle. These studies offer valuable insights on the state of species across landscapes and through time, demonstrating the DOM's suitability for monitoring oriented projects; including those where multiple species need to be assessed at once. Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships or taking a broader tack to identify drivers of occupancy without preconception. The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models. Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl. Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions as in @miller2023's study on American Kestrel nest boxes. In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as illustrated in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy. In recent years, increasing numbers of articles have used to DOM to generate predictions of species occupancy. As examples, @mcgowan2020 provides projections of future occupancy for the threatened Black rail under multiple management scenarios, and @pollentier2021 generates maps of distributions resembling those made with SDMs, outputs which can immediately be used by managers to inform action.

![A) Shows the proportion of articles in each year-strata and across all years which match each of six non-exclusive objective categories. B) Shows the quantity of covariates considered per parameter for models which pursued each objectives. Studies focused on methods development typically considered the fewest covariates, where those attempting to identify the drivers of occupancy considered the most.](Figures/ObjectiveFigure.jpeg){#fig-objectives}

The objectives of studies fitting a DOM strongly informs approaches to model building, particularly with respect to covariate selection. @fig-objectives B shows the quantity of covariates considered for models applied to each of our objective categories, and the differences between objectives are apparent. Not unexpectedly, articles which focused on describing new methods for DOMs had fewer covariates than those studies focused on more applied objectives. However, differences persist between those articles observing trends, identifying trends in occupancy, testing relationships, and making predictions.

@@

Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder. This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process. One aspect of fitting DOMs meriting broader discussion revolves around 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases. Complexity is a broad term which encompasses many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse. Where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014]. By their nature the DOM is somewhat more complex than simpler models for studying occupancy thanks to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics. Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters. Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for the DOM. Within the DOM framework, there are promising developments on that front — for example, @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.

Covariate selection seems to be a particularly important area for further investigations into building DOMs. In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models. This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method is most appropriate for their use-cases. In a similar vein, the low number of articles in our review that conducted model evaluation or assessed model fit raises different questions. While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is generally important to understand how reliable model outputs may be [@araújo2005; @guisan2005]. Existing uncertainties around whether current methods are suitable for the task likely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.

>>>>

abstract: | 
 Understanding the patterns by which species occur across landscapes and throughout time is a long-standing objective of ecological research which has inspired the development of numerous quantitative modelling approaches for the task. However, estimating occupancy can be a challenge, particularly where issues like imperfect detection and shifting distributions exist. Dynamic occupancy models (DOMs) offer a framework for occupancy estimation which explicitly accounts for observation error while incorporating additional mechanistic process by estimating colonisation and local extinction. In light of increasing interest in more process-explicit models for understanding species occurrence, here we examine how DOMs have been applied to field ecological data in the two decades since their introduction. Following a general introduction to the model, we present the results of a systematic review exploring the research contexts where DOMs have been applied and interrogating how authors have built their models, with particular emphasis on how covariates are incorporated to describe variation in occupancy dynamics. Our findings indicate that DOMs are a flexible tool readily applied to diverse study systems and data types, with their usage expanding in recent years as more studies apply them to make spatial and temporal prediction of species occupancy. DOMs are also amenable to extension, further broadening their utility. Despite their flexibility, model complexity in DOMs tends to be low; most studies consider relatively few covariates, most of which are represented as simple linear relationships. Approaches to covariate selection also vary considerably, and there remains little research on how these choices may influence model performance. Furthermore, only a fraction of articles report evaluating models and little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.

format:
  html:
    theme: simplex

embed-resources: true
title-block-banner: true
---

@@

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we recorded all covariates considered for each model in our sample, including those not used in final models. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of studies considering each type of covariate in their models, the means by which covariate data are collected, and how covariates responses are represented in DOMs. We further delineate their use in the model by classifying covariates into two groups: environmental covariates, representing plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that DOM users have incorporated a wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (35% of studies). Often these are included as simple covariates, as in @duggan2011's models using landcsape connectivety metrics as covariates on colonisation and extinction. Alternatively, more complex parameterisations explicitly model colonisation or extinction as a spatial process, blending attributes of DOMs and stochastic patch occupancy models [@broms2016; @risk2011]. Several studies also include biotic interactions with other species as covariates, particularly those focused on taxa threatened by invasive species. This is observed frequently in the Spotted owl literature, where Barred Owl presence is often considered as a covariate driving Spotted owl occupancy dynamics [@olson2005; ; @sovern2014; @dugger2016]. The use of these types of covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.

::: column-page-inset
```{r covariate table}
#| echo: false
#| message: false
#| label: tbl-covariates
#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories. We calculate the percentage of studies which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;). We also present the average percentage of covariates in a category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate with a non-linear relationship or as part of an interaction with another covariate."

@@

Covariate data for studies in our sample was either collected directly by researchers (an average of 30% of environmental covariates per study), or derived from pre-existing remotely sensed datasets (70% of environmental covariates); this of course varies depending on the category of covariate (@tbl-covariates). Directly collected data can often represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy. These species-specific covariates do come with trade-offs, as they can be expensive to collect and often preclude projection to locations where these data are unavailable. For studies interested in making such projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002]. An average of 43% percent of environmental factors and 94% of structural factors included in reviewed studies were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static (@tbl-covariates).

In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function) [@mackenzie2017]. Statistical relationships between model parameters and covariates (e.g., between initial occupancy and its environmental covariates) are represented as linear terms unless more complexity is specified. Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model. @austin2007 discusses the importance of modelling ecologically realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships. Non-linear responses can be easily accommodated in DOMs by using polynomial transformations and interactions between covariates. Despite this, in our sample only 35% of articles tested one or more non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear terms (@tbl-covariates). Interactions between covariates were similarly rare, with only 24% of studies considering at least one interaction between terms (@tbl-covariates). The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy. Many common approaches for SDMs, such as MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves and use of interactions where supported by the data [@merow2013; @elith2008]. This emphasis on more complex responses in SDMs may be due to their frequent application across relatively large geographic extents which might encompass the full species niche, where environmental relationships may be expected to be non-linear. However, as previously indicated DOMs have also been implemented across large spatial extents where the same assumptions should exist and similar responses might be expected. Similar concerns exist for the relatively low level of covariate interactions, given that these relationships are commonly expected based on ecological theory and that their exclusion can negatively impact model performance [@guisan2006].

In addition to the types of covariates considered for modelling we also recorded the size of the candidate covariate pool in each reviewed study, tallying the number of environmental and structural covariates which were available for use on each parameter ( @fig-covariates). This is an area of considerable variation in modelling practices — the number of covariates considered for each parameter ranges from 0 (effectively modelling the parameter as a constant) to over 40 candidates on a single parameter.

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents the alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

The median number of covariates varies strongly by parameter, with transition probabilities (colonisation and extinction) more likely to consider a broader range of environmental covariates compared to initial occupancy and detection. The lack of any covariates considered for initial occupancy in many studies (37% of studies) is particularly notable — unless a study is conducted at very small extents or study sites are truly uniform in their suitability, one would expect some amount of non-random variation in occupancy probability across any study system that will not be captured when representing initial occupancy as a constant. Omission of the factors that drive occupancy is a known source of bias in static SDMs [@barry2006], which the initial occupancy component of the DOM conceptually resembles. Furthermore, any bias in occupancy estimation in the first time-step will perpetuate into future seasons due to the DOM's Markovian nature with implications for the reliability of model outputs. Detection probability is also typically represented with low numbers of environmental covariates (@tbl-covariates). Recall that detectability is dependent on not only observation factors, but also by drivers of abundance and species use; the low number of candidate covariates here relative to colonisation and extinction raises additional questions on whether this variation is always appropriately captured in models.

To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference. Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012]. Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$: where a simple linear regression may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations of these same covariates. This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires some reduction of the models tested. Models in our sample include both Frequentist (n = 76) and Bayesian (n = 24) frameworks, which differ considerably in their manner of implementation; some articles in our sample included multiple distinct modelling workflows, resulting in a higher total number of models than articles. A summary of modelling practices in DOMs is given in @tbl-modelling.

```{r Modelling summary table}
#| echo: false
#| message: false
#| label: tbl-modelling
#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by frequentist or bayesian implementations. The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review. The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches. 2 models included in the 'Overall' column are neural network based and fall into neither the Frequentist or Bayesian categories."

@@

The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection. Where 95% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 33% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*. Methods used for model selection vary considerably both within and between the two modelling frameworks (@tbl-modelling). For frequentist models, the most popular and conventional approach to model selection (45% of models) involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score. The next most popular method in frequentist studies is to use procedural model selection methods (37% of models), where the structure for each model parameter is fit in sequence. For example, this protocol might first identify the best structure for detection probability while holding the other parameters constant, before moving on to initial occupancy and so on until all parameters are fixed. The remainder of frequentist studies (8%) use a variety of approaches, such as fitting simpler models like single season occupancy models to identify the most informative terms to use in a DOM. Across all frequentist implementations, 47% of articles summarise a final subset of models using multi-model inference and model-averaging with AIC weights [@burnham2004].

Those Bayesian models which *do* perform model selection take various approaches, with largely idiosyncratic methods across these studies. While direct comparison of model fit is rare in Bayesian methods, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data. Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model. @ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure as a Bayesian model.

Limited research has been conducted on the advantages of different methods for covariate selection in DOMs, and there is unlikely to be a one-size-fits-all approach which will be appropriate for all possible use-cases. However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020]. Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy models discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate coefficient estimates that may influence model inference. @morin2020's work raises other concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models. The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.

Regardless of covariate selection protocol, 'the selection of a best model does not guarantee the selection of a good model' [@mackenzie2004]. Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not yet clear. There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004's approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages. @kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model. In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014]. @broms2016 discusses further possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models, and extensions of their approach may also be applicable to the DOM. As with other hierarchical models, model evaluation for DOMs can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known. Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.

Perhaps as a result of these difficulties, assessment of model fit and model performance was rare amongst the articles in our sample. Only 18% of studies tested for goodness-of-fit, and just 7% calculated predictive performance. These rates are considerably lower than those reported for SDMs; for which closer to 50% of articles were found to test both fit and performances [@araújo2019].

@@

Studies in our sample were classified according to their objectives as stated by authors to explore the use-cases which the DOM has been applied to (@fig-objectives). These objective categories are non-exclusive, with many studies falling into multiple categories. Across the study period the DOM has been frequently employed to estimate trends in species occupancy, to explore relationships between environmental factors and occupancy, and, increasingly often, to make predictions of occupancy spatially to unsurveyed locations or temporally to hypothetical future conditions. Within each of these broad categories lies even more variation in objectives, emphasising the DOM's flexibility as a tool for making ecological inference in diverse contexts.

36% of articles used the DOM to monitor trends in occupancy state through time ("Estimate trends" in @fig-objectives A), both for single species of high conservation interest and for broader community assemblages of species [@ahumada2013; @scott2015]. These studies offer valuable insights on the state of species across landscapes and through time, demonstrating the DOM's suitability for monitoring oriented projects; including those where multiple species need to be assessed at once. Other articles were more focused on examining the factors which influence species occupancy, either testing pre-specified hypothetical relationships (44% of articles) or taking a broader tack to identify drivers of occupancy without preconception (35%). The distinction between the two is important, as it guides which covariates may be considered and how model selection may be used to fit models. Many of these studies target core conservation priorities for their focal species, like @olson2005's early DOM assessing the influence of barred owls on threatened spotted owl. Explorations of these pivotal relationships are important for guiding management action, and may also be used to test the effectiveness of these actions [@miller2023]. In scenarios where less is known about species habitat preference, DOMs may be used to examine the influence of a wider variety of factors as in @huber2017's study testing the relative influence of dozens of habitat covariates on Wood warbler occupancy. In recent years, increasing numbers of articles have used DOMs to generate predictions of species occupancy; this a use-case for which DOMs show considerable promise [@kéry2013; @briscoe2021]. These predictions often have strong utility for conservation management. For example, @mcgowan2020 projects future occupancy for wetland birds under alternative management scenarios, and @pollentier2021 generates distribution maps resembling those made with SDMs.

![A) Proportion of articles in each year-strata and across all years which match each of six non-exclusive objective categories. B) Quantity of covariates considered per parameter for models which pursued each objectives.](Figures/ObjectiveFigure.jpeg){#fig-objectives}

Study objectives strongly inform approaches to model building, particularly with respect to covariate selection. @fig-objectives B presents the quantity of covariates considered for models applied to each of our objective categories, with apparent differences between objectives. Not unexpectedly, articles which focused on describing new methods for DOMs had fewer covariates than those studies focused on more applied objectives. However, differences persist between those articles observing trends, identifying trends in occupancy, testing relationships, and making predictions.

@@

Approaches to building any type of model will necessarily depend on the possibilities of the data at hand and on the priorities of the model-builder. This precludes any prescription of the 'best' way to build a model, however, there are still important discussions to be had on decisions made in the modelling process. One aspect of fitting DOMs meriting broader discussion centres on 'model complexity,' and how much must be incorporated into models to reliably model species occupancy under different contexts and use cases. Complexity is a broad term encompassing many aspects of a model [@merow2014], and opinions on simplicity versus complexity in ecological models are diverse. Where some advocate for the simplest possible models, arguing that they are most generalisable; others insist that overly-simple models cannot adequately represent the most important drivers in a system [@evans2013; @lonergan2014]. By their nature the DOM is somewhat more complex than simpler models for studying occupancy due to their hierarchical structure, which is necessary to control for detectability and to capture occupancy dynamics. Within this structure, however, further complexity is to some degree up to the modeller: one can choose how many covariates to consider for inclusion on the various parameters, and how to represent the nature of the relationship between those covariates and parameters. Research from SDMs indicates that allowing for more complex relationships can improve model performance in predicting occupancy [@valavi2023], an increasingly popular use-case for the DOM. Within the DOM framework, there are promising developments on that front: @joseph2020 presents a novel neural-network occupancy model which allows for exponentially higher levels of complexity and may offer improved performance for prediction-oriented studies.

Covariate selection seems to be a particularly important area for further investigations into building DOMs. In our review, we see little consensus around which approaches are most applicable for any given use case, and existing work on covariate selection in DOMs raises concerns on whether common methods always produce the most suitable models [@morin2020; @stewart2023]. This is true for both frequentist and Bayesian implementations, and comparative research on covariate selection under both frameworks may help to inform model users on which method is most appropriate for their use-cases. In a similar vein, the low number of articles in our review that conducted model evaluation or assessed model fit raises different questions. While the appropriate method of model evaluation may depend on data availability and research objectives, assessing models by some method is generally important to understand how reliable model outputs may be [@araújo2005; @guisan2005]. Existing uncertainties around whether current methods are suitable for the task likely discourage users from calculating these metrics, and further research is needed to establish trusted practices for assessing the quality of DOMs.

These priorities are particularly important given the frequent applied objectives of DOM users, tackling challenges which include assessing critically endangered species, guiding public health management of disease vectors, and tracking rapidly developing biological invasions [@carvalho2023; @moreira2016; @wood2020]. The DOM is well suited to these situations, and it is to be expected that as these types of applications are more commonly attempted understanding the sensitivity of model outputs to decisions made in the model fitting process becomes increasingly important. In the two decades since the publication of @mackenzie2003 the dynamic occupancy model has become a widely used tool for ecological inference, with numerous extensions to the modelling framework further broadening the scope of questions and use-cases for which it may be applied. Given their increasing popularity, further research and guidelines around issues of model building may help to make the DOM more accessible to newcomers and support confidence in model interpretation. Parallels to existing work in the SDM literature on guidelines for reporting and modelling standards could be valuable contributions in achieving these aims and provide useful examples in developing methods and guidelines specifically focused on DOMs [@zurell2020; @araújo2019].
